Rapid Epidemiological Analysis of Comorbidities and Treatments in COVID-19 in Scotland (REACT-SCOT): a case-control study based on the Scottish population [**Draft, not to be quoted**]

### 2020-05-14 casecontrol.Rmd
updated for ethnicity report

---
title: 'Rapid Epidemiological Analysis of Comorbidities and Treatments in COVID-19 in Scotland (REACT-SCOT): a case-control study based on the Scottish population'
author: 
  - name: Paul M McKeigue
    email: paul.mckeigue@ed.ac.uk
    affiliation: Usher
  - name: Amanda Weir 
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Stuart McGurnaghan
    affiliation: IGMM
  - name: Sharon Kennedy
    affiliation: HPS
  - name: David McAllister 
    affiliation: Glasgow
  - name: Chris Robertson 
    affiliation: Strathclyde
  - name: Rachael Wood
    affiliation: ISD
  - name: Nazir Lone
    affiliation: Usher
  - name: Janet Murray
    affiliation: HPS 
  - name: Tom Caparrotta 
    affiliation: IGMM
  - name: Sharon Hutchinson 
    affiliation: GCU
  - name: Helen M Colhoun
    affiliation: IGMM
address: 
   - code: Usher
     address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care
   - code: IGMM
     address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - Axa Chair in Medical Informatics and Epidemiology. TC - Sir George Alberti Doctoral Fellow in Pharmacoepidemiology. 
   - code: Glasgow
     address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
   - code: HPS
     address: Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
   - code: Strathclyde
     address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
   - code: GCU
     address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
   - code: ISD
     address: NHS Information Services Division (Public Health Scotland), Gyle Square, 1 South Gyle Crescent, Edinburgh, EH12 9EB.  RW - Consultant in Maternal and Child Health. 
geometry: margin=1in
fontsize: 11pt
linenumbers: true
bibliography: ../covid.bib
date: "This version compiled `r gsub('^0', '', format(Sys.time(), '%d %B %Y'))`"
header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{lscape}
output: 
#  pdf_document:
  rticles::plos_article:
    latex_engine: lualatex
    fig_caption: true
    keep_tex: true
#    toc: true
#    toc_depth: 2
csl: ../plos.csl
urlcolor: red

@@

```{r functions, echo=FALSE}

@@

library(knitr)
library(kableExtra)

@@

_Background_-- Why some people infected with Sars-CoV-2 develop severe COVID-19 disease and others do not is poorly understood.  Here we  used sociodemographic data and extensive previous hospitalisation and prescribing history among all people who have developed severe COVID-19 disease in Scotland compared to the background population to understand factors associated with severity.

_Methods_ -- We linked the records of those tested positive for SARS-C0V-2 viral RNA in the national database that captures virology testing in all NHS laboratories, to  hospital admissions, intensive care and mortality data. Severe COVID-19 was defined as disease requiring intensive care support or associated with death within 28 days of a positive test. We selected 10 sex, age and general practice based controls per test-positive case. For all cases and controls we extracted the International Classification of Disease (ICD)-10 codes from the previous five years of hospital discharge records  and the British National Formulary (BNF)  drug codes of  encashed  primary care issued prescriptions in the past year. We evaluated the risk of severe COVID-19 associated with  a pre-defined  set of  conditions using conditional logistic regression. We examined how much predictive performance  for severe COVID-19  a risk classifier built  and tested with four-fold cross validation based on these conditions could achieve and whether this could be improved using information from the rest of the hospitalisation and prescribing data.

_Findings_ -- In a logistic regression based on the age-sex distribution of the national population, the odds ratios were `r round(exp(logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.  In the case-control analysis, the odds ratios were 
`r table.agegr.all[1, 3]` for residence in a care home status, 
`r table.agegr.all[4, 3]` for diabetes, 
`r table.agegr.all[8, 3]` for chronic renal failure, 
`r table.agegr.all[5, 3]` for ischemic heart disease
`r table.agegr.all[9, 3]` for neurological disease, 
`r table.agegr.all[10, 3]` for chronic liver disease and 
`r table.agegr.all[11, 3]` for immune deficiency or suppresssion

Overall `r freqs.all[13, 2]` of severe cases and `r freqs.all[13, 2]` of controls had at least one of these listed conditions: `r gsub("\\\\", "", table.agegr[13, 2])` versus `r gsub("\\\\", "", table.agegr[13, 1])` in those aged under 40 years).  The odds ratio for severe disease associated with encashment of at least one prescription drug in the past year was `r table.agegr.all[2, 3]` and the odds ratio associated with at least one hospital admission in the past five years was `r table.agegr.all[3, 3]`).  These associations with severe disease were present in those without any of the listed conditions, and were distributed over most BNF chapters and the associations with hospital diagnoses over most ICD-10 chapters.

With a model based on age and sex alone, the expected information for discrimination was `r round(cases.Lambda.agesex, 2)` bits.  The additional information for discrimination conditional on age and sex was `r table.densities[1, 5]` bits for a model based on demographic variables only, `r table.densities[2, 5]` bits for a model that included listed condtions, and `r table.densities[3, 5]` bits for a model that included indicator variables for hospital diagnoses and prescription encashments.

_Conclusions_ --Along with older age and male sex, severe COVID-19 disease is strongly associated with past medical history and most younger persons developing severe disease have an underlying medical condition. The comorbidities associated with increased risk are not limited to the current lists of risk conditions highlighted by public health agencies.  A risk classifier that uses all the information available in medical and prescribing records, rather than only a limited set of conditions, will more accurately discriminate between low-risk and high-risk individuals who may require shielding until the SARS-CoV-2 epidemic is over.   The strong association with prescribing history we have observed may reflect co-morbidity or adverse drug effects and we are investigating this further.

@@

# Background
It is clear that  there is wide susceptibility to risk of severe COVID-19 disease  with older age being a predominant factor. Case series from many countries have  suggested that in those with severe disease the prevalence of diabetes and cardiovascular disease is higher than expected.  However there are also many anecdotal reports of apparently healthy young persons succumbing to disease.  Apart from co-morbidities the potential  role of ethnicity in disease severity has been highlighted.  However quantification of the risk associated with these characteristics and co-morbidities has been limited by the lack of an appropriate background population comparator.  In particular a systematic evaluation of the past medical history in those with and without severe disease has not been reported.

Here we  used extensive health care record and prescribing history among all people who have developed severe COVID-19 disease in Scotland compared to the background population to understand sociodemographic and comorbidity factors associated with severe COVID -19 disease.  We defined  severe disease as viral RNA-confirmed diagnosis of COVID-19 as  that required critical care support or was associated with death within 28 days of testing.   We focused on severe disease it is of most  concern but also  because it can be completely ascertained  whereas  testing policies means that   complete ascertainment of less severe disease is not possible in most countries.

@@

## Matched controls
For each test-positive case, we ascertained ten matched controls of the same sex, one year age band and attending the same General Practice as each case alive on the same day as the date of test in the case using the Community Health Index (CHI) database. As this is an incidence density sampling design, it is possible and correct for an individual to appear in the dataset more than once, initially as a control and subsequently as a case.  With this design, odds ratios in conditional logistic regression models are equivalent to rate ratios.

@@

We first extracted history of a specific list of conditions and drug classes that have been included as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].  These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an encashed prescription of a drug for which that diagnostic code is the only indication.  The ICD-10 codes and drug codes used to define each variable are listed in the R script provided as Supplementary Material. In addition to these pre-specified groupings of codes we then undertook a systematic evaluation of the association of all ICD-10 codes and BNF codes to chapter level.  We did not specifically tabulate the conditions that have been listed as making people " clinically extremely vulnerable" such as cystic fibrosis or leukaemia as there is no clinical equipoise regarding risk in such persons and as it is expected that they will be underrepresented in the case data as they have been advised to shield themselves completely since early in the epidemic.

## Statistical methods
To estimate the relation of cumulative incidence and mortality from COVID-19 to age and sex, logistic regression models were fitted to the proportions of cases and non-cases in the Scottish population, using the estimated population of Scotland in mid-year 2019 which are available by one-year age group up to age 90 years. For the case-control study, estimates of associations with severe COVID-19 were based on conditional logistic regression, implemented as Cox regression in the R function `survival::clogit`.  Among those cases and controls without any of the pre-defined conditions we then further examined associations of ICD-10 and BNF chapter with severe COVID-19. To construct risk prediction models, we used stepwise regression alternating between forward and backward steps to maximize the AIC, implemented in the R function `stats::step`.  The performance of the risk prediction model as a classifier was examined by 4-fold cross-validation, with performance calculated over all test folds as the expected information for discrimination [@mckeigue_quantifying_2019] and (more conventionally) as the C-statistic.

## Information governance
This study was conducted under approvals from the Public Benefit and Privacy Panel that allow Health Protection Scotland staff to link any PHI controlled datasets <!--- janet to add detail and reference numbers ---> and amendment 1617-0147 that covers transfer of diabetes data to Health Protection Scotland. Datasets were de-identified before analysis.

# Results
<!---
Still needed 
- A statement on "In national mortality data, the n and % of those certified with COVID-19 as underlying case of death had/ did not have a positive test record"  or sensivitivty analysis if sharon provides a case control set of the deaths with ue07 code 
- Figure - ?? graph of population attributable risk fraction of listed conditions by age ?? not sure if this is worth the trouble ...
--->
## Incidence and mortality from severe COVID-19 in the Scottish population
Figure \ref{fig:rates} shows the relationships of incidence and mortality rates to age for each sex separately.  The relationship of mortality to age was almost exactly linear on a logit scale, and the lines for males and females were almost exactly parallel.  For severe cases as defined in this study, the incidence appears to rise more slowly from age 50 to 70 years in females than in males, implying that the ratio of non-fatal cases who receive critical care to fatal cases is lower in females than in males in this age range.  This may reflect either an underlying divergence in case fatality, or that women are less likely than men to be transferred to critical care if they have a good chance of surviving without critical care.  In models including both age and sex as covariates, the odds ratio associated with a 10-year increase in age was `r round(exp(logistic.coeffs[3, 1]), 2)` for all severe disease and `r round(exp(logistic.coeffs[3, 2]), 2)` for fatal disease.  The odds ratio associated with male sex was `r round(exp(-logistic.coeffs[2, 1]), 2)` for all severe disease and `r round(exp(-logistic.coeffs[2, 2]), 2)` for fatal disease.  As shown in Table \ref{tab:agegr} almost all severe cases over aged 75 years died but most severe cases under aged 60 survived.

@@

## Associations with listed conditions
Overall `r freqs.all[13, 2]` of severe cases and `r freqs.all[13, 1]` of controls had at least one of the listed conditions.  The prevalence of the listed conditions in cases and controls by age band is shown in Table \ref{tab:agegr}.  `r gsub("\\\\", "", table.agegr[13, 2])` of the cases aged under 40 years had at least one listed condition, compared with only `r gsub("\\\\", "", table.agegr[13, 1])` of the controls.  In those aged 75+ years `r gsub("\\\\", "", table.agegr[13, 6])` of the cases and `r gsub("\\\\", "", table.agegr[13, 5])` of the controls had at least one listed condition.

@@

## Systematic analysis of diagnoses associated with severe disease
The approximately two-fold risk of with severe COVID-19 associated with prior hospital admission independently of other conditions and factors was further explored. The association of hospital diagnoses of each ICD-10 chapter with severe COVID-19, restricted to those without any of the listed conditions.  These results are shown in Table \ref{tab:conditions}.  In univariate analyses, almost all ICD chapters with at least one diagnosis, with the exception of Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.  Table \ref{tab:icdsubchapters} extracts univariate associations with ICD subchapters: .. note associations with previous pneumonia and stroke.

## Associations of prescribed drugs with severe disease 
Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that the strongest risk factor for severe disease, apart from residence in a care home, is the encashment of any prescription in the last year.  The rate ratio associated with this variable varies from `r tables.agegr[[1]][2, 3]` in those aged under 60 years to `r tables.agegr[[3]][2, 3]` in those aged 75 years and over.  In a multivariate analysis adjusting for care home residence, any hospital admission and listed conditions, these rate ratios were reduced to `r tables.agegr[[1]][2, 5]` and  `r tables.agegr[[3]][2, 5]`   About one third of those over 75 had not encashed a prescription in the previous year.

@@

## Construction of a multivariate risk prediction model 
Predictive models were constructed from two sets of variables: a baseline set consisting only of care home residence plus the listed conditions, and an extended set that augmented the baseline set with an indicator variable for ICD-10 chapter level admissions, and indicator variables for prescriptions in each of the BNF chapters.

@@

# Discussion  <!--helen still to edit -->
## Multifactorial risk
In this analysis of sociodemographic factors and past medical history we found a large number of factors to contribute to risk of severe COVID-19 disease.  Apart from the very steep rise in risk with age, severe disease is more likely in men, and those who are more socioeconomically disadvantaged. Being in a residential care home was associated with a 15-fold increased risk of severe COVID-19 in this age matched analysis.

In contrast reports from the U.S and elsewhere in the UK [@the_opensafely_collaborative_opensafely_2020] we found no evidence of an increased risk in non-White minorities. This suggests that ethnic differences found elsewhere are not genetically driven but reflect other social factors. In a recent report based on the UK Biobank cohort, almost half of the fourfold risk ratio for hospitalised COVID associated with Black ethnicity was accounted for by the health care worker status [@niedzwiedz_ethnic_2020;@ho_modifiable_2020 ].

Media reports of apparently otherwise healthy youths succumbing to severe COVID-19 can lead to an impression that severe COVID-19 is a random event. However we found that almost half of cases under 40 years had at least one condition listed by public health agencies as increasing risk.  In those that did not have a listed condition  having any hospital admission  in the past five years and having any prescription in the past year  was much more common  among cases that in the background population of the same age.

We chose to focus initially on the contribution of  conditions already proposed as high risk.  We confirmed that these are associated with  substantially increased risk .  However our systematic analysis has shown that there are many many  many other conditions considered in the same univariate way are also  associated with severe COVID-19 disease.

## Possible drug effects
A striking finding of this study was the very strong association of severe COVID-19 with having encashed at least one prescription in the past year. This association was only partly explained by associations with listed conditions.  Partitioning of this association between BNF chapters, which represent broad indication-based drug classes, showed that the strongest association was with prescription of Chapter 1 drugs, prescribed for gastrointestinal conditions, which are not generally listed as risk factors for severe COVID-19.  In an accompanying paper we explore these associations more fully.

## Methodological strengths and weaknesses
Most of the reports of disease associations with COVID-19 have been case series descriptions there are few prospective cohort or case control analyses with appropriate population comparators.  With this matched case control design using incidence density sampling, we have been able to estimate rate ratios conditional on age and sex.  An unpublished analysis from England explored the association of similar set of risk conditions with in-hospital COVID-19 deaths, but did not systematically evaluate the rest of the medical record including prescription records.  Although we have records of encashment of prescriptions, we do not at present have access to other the primary care data, which would contain additional information such as body mass index.  Hospital discharge diagnoses however are coded to ICD-10 by trained coders.  A further limitation is generalisability especially of sociodemographic factors to other locations. Also we limited our case control study to test-positive cases. This excludes deaths attributed to COVID-19 where no testing was carried out. At present there are xxx such deaths in Scotland almost all aged >> xxx.  Others have argued that some COVID-19 deaths may have been attributed to other causes early in the epidemic.  These aspects may mean some relative risks e,g that for residential care home status have been underestimated.  However in the younger age bands, it is likely that most severe or fatal cases will have reached hospital and been tested.

>>>>

---
title: 'Rapid Epidemiological Analysis of Comorbidities and Treatments in COVID-19 in Scotland (REACT-SCOT): a case-control study based on the Scottish population [**Draft, not to be quoted**]'
author: 
  - name: Paul M McKeigue
    email: paul.mckeigue@ed.ac.uk
    affiliation: Usher
  - name: Amanda Weir 
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Stuart McGurnaghan
    affiliation: IGMM
  - name: Sharon Kennedy
    affiliation: HPS
  - name: David McAllister 
    affiliation: Glasgow
  - name: Chris Robertson 
    affiliation: Strathclyde
  - name: Rachael Wood
    affiliation: ISD
  - name: Nazir Lone
    affiliation: Usher
  - name: Janet Murray
    affiliation: HPS 
  - name: Tom Caparrotta 
    affiliation: IGMM
  - name: Sharon Hutchinson 
    affiliation: GCU
  - name: Helen M Colhoun
    affiliation: IGMM
address: 
   - code: Usher
     address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care
   - code: IGMM
     address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - Axa Chair in Medical Informatics and Epidemiology. TC - Sir George Alberti Doctoral Fellow in Pharmacoepidemiology. 
   - code: Glasgow
     address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
   - code: HPS
     address: Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
   - code: Strathclyde
     address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
   - code: GCU
     address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
   - code: ISD
     address: NHS Information Services Division (Public Health Scotland), Gyle Square, 1 South Gyle Crescent, Edinburgh, EH12 9EB.  RW - Consultant in Maternal and Child Health. 
geometry: margin=1in
fontsize: 11pt
linenumbers: true
bibliography: ../covid.bib
date: "This version compiled `r gsub('^0', '', format(Sys.time(), '%d %B %Y'))`"
header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{lscape}
output: 
#  pdf_document:
  rticles::plos_article:
    latex_engine: lualatex
    fig_caption: true
    keep_tex: true
#    toc: true
#    toc_depth: 2
csl: ../plos.csl
urlcolor: red

@@

```{r functions, echo=FALSE}
library(knitr)
library(kableExtra)

@@

@@

_Background_-- Why some people infected with SARS-CoV-2 develop severe COVID-19 disease and others do not is poorly understood.  Here we  used sociodemographic data and extensive previous hospitalisation and prescribing history among all people who have developed severe COVID-19 disease in Scotland compared to the background population to understand factors associated with severity.

_Methods_ -- We linked the records of those with a positive nucleic acid test  for SARS-CoV-2 in the national database that captures virology testing in all NHS laboratories, to  hospital admissions, intensive care and mortality data. Severe COVID-19 was defined as disease requiring intensive care support or associated with death within 28 days of a positive test. We selected 10 sex, age and general practice based controls per test-positive case. For all cases and controls we extracted the International Classification of Disease (ICD)-10 codes from the previous five years of hospital discharge records  and the British National Formulary (BNF)  drug codes of  encashed  primary care issued prescriptions in the past year. We evaluated the risk of severe COVID-19 associated with  a pre-defined  set of  conditions using conditional logistic regression. We examined how much predictive performance  for severe COVID-19  a risk classifier built  and tested with four-fold cross validation based on these conditions could achieve and whether this could be improved using information from the rest of the hospitalisation and prescribing data.

_Findings_ -- In a logistic regression based on the age-sex distribution of the national population, the odds ratios were `r round(exp(10 * logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.  In the case-control analysis, the odds ratios were 
`r table.agegr.all[1, 3]` for residence in a care home status, 
`r table.agegr.all[4, 3]` for diabetes, 
`r table.agegr.all[8, 3]` for chronic renal failure, 
`r table.agegr.all[5, 3]` for ischemic heart disease,
`r table.agegr.all[6, 3]` for other heart disease, 
`r table.agegr.all[9, 3]` for neurological disease, 
`r table.agegr.all[10, 3]` for chronic liver disease and 
`r table.agegr.all[11, 3]` for immune deficiency or suppresssion.

Overall `r freqs.all[13, 2]` of severe cases and `r freqs.all[13, 1]` of controls had at least one of these listed conditions: `r gsub("\\\\", "", table.agegr[13, 2])` versus `r gsub("\\\\", "", table.agegr[13, 1])` in those aged under 40 years.  The odds ratio for severe disease associated with encashment of at least one prescription drug in the past year was `r table.agegr.all[2, 3]` and the odds ratio associated with at least one hospital admission in the past five years was `r table.agegr.all[3, 3]`).  These associations with severe disease were present in those without any of the listed conditions, and were distributed over most BNF chapters and the associations with hospital diagnoses over most ICD-10 chapters.

With a model based on age and sex alone, the expected information for discrimination was `r round(cases.Lambda.agesex, 2)` bits.  The additional information for discrimination conditional on age and sex was `r table.densities[1, 5]` bits for a model based on demographic variables only, `r table.densities[2, 5]` bits for a model that included listed conditions, and `r table.densities[3, 5]` bits for a model that included indicator variables for hospital diagnoses and prescription encashments.

_Conclusions_ --Along with older age and male sex, severe COVID-19 disease is strongly associated with past medical history and most younger persons developing severe disease have an underlying medical condition. The comorbidities associated with increased risk are not limited to the current lists of risk conditions highlighted by public health agencies.  A risk classifier that uses all the information available in medical and prescribing records, rather than only a limited set of conditions, will more accurately discriminate between low-risk and high-risk individuals who may require shielding until the SARS-CoV-2 epidemic is over.

@@

# Background
Case series from many countries have suggested that in those with severe COVID-19 the prevalence of diabetes and cardiovascular disease is higher than expected.  For example in a large UK series the commonest co-morbidities were cardiac disease, diabetes, chronic pulmonary disease and asthma [@docherty_features_2020]. However there are also anecdotal reports of apparently healthy young persons succumbing to disease [@mcgoogan_why_2020].

The quantification of the risk associated with  characteristics and co-morbidities has been limited by the lack of an appropriate background population comparator  [@zhou_clinical_2020;@grasselli_baseline_2020;@guan_clinical_2020].   Two recent UK analyses have  had population comparators and these have reported  increased relative risks for a range of co-morbidities including diabetes, asthma and heart disease and for non-white ethnicity[@niedzwiedz_ethnic_2020;@the_opensafely_collaborative_opensafely_2020].   A systematic evaluation of the past medical history in those with and without severe disease has not as yet been reported.

We undertook a systematic examination of the health care record and prescribing history among all people who have developed severe COVID-19 disease in Scotland compared to the background population to understand sociodemographic and comorbidity factors associated with severe COVID-19 disease.

@@

## Matched controls
For each test-positive case, we ascertained ten matched controls of the same sex, one year age band and attending the same General Practice as each case alive on the same day as the date of test in the case using the Community Health Index (CHI) database. As this is an incidence density sampling design, it is possible and correct for an individual to appear in the dataset more than once, initially as a control and subsequently as a case.  With this design, odds ratios in conditional logistic regression models are equivalent to rate ratios.  The odds ratios based on the conditional likelihood will differ from the unconditional odds ratios calculable from the frequencies of exposure in cases and controls [@breslow_estimation_1978]

@@

We first extracted history of a specific list of conditions and drug classes that have been included as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].  These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an encashed prescription of a drug for which that diagnostic code is the only indication.  The ICD-10 codes and drug codes used to define each variable are listed in the R script provided as Supplementary Material. In addition to these pre-specified groupings of codes we then undertook a systematic evaluation of the association of all ICD-10 codes and BNF codes to chapter level.  We did not specifically tabulate the conditions that have been listed as making people " clinically extremely vulnerable" such as cystic fibrosis or leukaemia  or organ transplantation as there is no clinical equipoise regarding risk in such persons and as it is expected that they will be underrepresented in the case data as they have been advised to shield themselves completely since early in the epidemic. This group also includes those with cancer in receipt of chemotherapy or radiotherapy.

## Statistical methods
To estimate the relation of cumulative incidence and mortality from COVID-19 to age and sex, logistic regression models were fitted to the proportions of cases and non-cases in the Scottish population, using the estimated population of Scotland in mid-year 2019 which are available by one-year age group up to age 90 years. For the case-control study, estimates of associations with severe COVID-19 were based on conditional logistic regression, implemented as Cox regression in the R function `survival::clogit`.  Among those cases and controls without any of the pre-defined conditions we then further examined associations of ICD-10 and BNF chapter with severe COVID-19. Where the sample of cases and controls is restricted, this will generate strata that contain no cases.  These strata will be ignored by the conditional logistic regression model as they do not contribute to the conditional likelihood.

To construct risk prediction models, we used stepwise regression alternating between forward and backward steps to maximize the AIC, implemented in the R function `stats::step`.  The performance of the risk prediction model as a classifier was examined by 4-fold cross-validation, with performance calculated over all test folds as the expected information for discrimination $\Lambda$ [@mckeigue_quantifying_2019] and (more conventionally) as the C-statistic. The rationale for using $\Lambda$ to quantify the predictive performance of a binary classifier has been explained elsewhere [@mckeigue_quantifying_2019].  Especially relevant to this study is that the contributions of independent predictors are additive on the scale of $\Lambda$.  We can thus add the information for discrimination conditional on age and sex (estimated from this matched case-control study) to the information for discrimination estimated from a logistic regression model on age and sex in the general population.

## Information governance
This study was conducted under approvals from the  Privacy Advisory Committee ref 44/13  and Public Benefit Privacy Protection amendment 1617-0147 . Datasets were de-identified before analysis.

# Results
## Incidence and mortality from severe COVID-19 in the Scottish population
Figure \ref{fig:rates} shows the relationships of incidence and mortality rates to age for each sex separately.  The relationship of mortality to age is almost exactly linear on a logit scale, and the lines for males and females are almost exactly parallel.  For severe cases as defined in this study, the sex differential is narrow up to about age 50, implying that at younger ages the ratio of critical care admissions to total fatalities is higher in women than in men, but widens from about age 40 onwards.  In models including both age and sex as covariates, the odds ratio associated with a 10-year increase in age was `r round(10*exp(logistic.coeffs[3, 1]), 2)` for all severe disease and `r round(exp(10*logistic.coeffs[3, 2]), 2)` for fatal disease.  The odds ratio associated with male sex was `r round(exp(-logistic.coeffs[2, 1]), 2)` for all severe disease and `r round(exp(-logistic.coeffs[2, 2]), 2)` for fatal disease.  As shown in Table \ref{tab:agegr} almost all severe cases over aged 75 years died but most severe cases under aged 60 survived.

@@

## Associations with listed conditions
Overall `r freqs.all[13, 2]` of severe cases and `r freqs.all[13, 1]` of controls had at least one of the listed conditions.  The prevalence of the listed conditions in cases and controls by age band is shown in Table \ref{tab:agegr}.  `r gsub("\\\\", "", table.agegr[13, 2])` of the cases aged under 40 years had at least one listed condition, compared with only `r gsub("\\\\", "", table.agegr[13, 1])` of the controls.  In those aged 75+ years `r gsub("\\\\", "", table.agegr[13, 8])` of the cases and `r gsub("\\\\", "", table.agegr[13, 7])` of the controls had at least one listed condition.

@@

## Systematic analysis of diagnoses associated with severe disease
The approximately two-fold risk of with severe COVID-19 associated with prior hospital admission independently of other conditions and factors was further explored by systematically testing for association of hospitalisations at each ICD-10 chapter level with severe COVID-19, restricted to those without any of the listed conditions.  These results are shown in Table \ref{tab:conditions}.  In univariate analyses, almost all ICD chapters, with the exception of Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.  The strongest associations were seen with neoplasms and with eye disease. Table \ref{tab:icdsubchapters} extracts univariate associations with ICD subchapters where the p-value is <0.001 and where there are at least 50 cases and controls with a diagnosis in this subchapter.  The key point of this table shows that there are many subchapter diagnoses that are significantly more frequent in severe COVID-19 cases than controls.  Of particular note is that prior hospitalisations with infections, pneumonia and acute respiratory diseases, circulatory disorders other than heart disease, including cerebrovascular accidents.  Without filtering on the aforementioned criteria there are of course additional differences in counts of rarer diagnoses in cases and controls are apparent.

## Associations of prescribed drugs with severe disease 
Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that the strongest risk factor for severe disease, apart from residence in a care home, is the encashment of any prescription in the last year.  The univariate rate ratio associated with this variable varies from `r tables.agegr[[1]][2, 3]` in those aged under 60 years to `r tables.agegr[[3]][2, 3]` in those aged 75 years and over.  In a multivariate analysis adjusting for care home residence, any hospital admission and listed conditions, these rate ratios were reduced to `r tables.agegr[[1]][2, 5]` and  `r tables.agegr[[3]][2, 5]`   About one third of those over 75 had not encashed a prescription in the previous year.

@@

## Construction of a multivariate risk prediction model 
To obtain a summary metric of the additional information there is in the hospitalisation and prescribing data predictive models were constructed from two sets of variables: a baseline set consisting only of care home residence plus the listed conditions, and an extended set that augmented the baseline set with an indicator variable for ICD-10 chapter level admissions, and indicator variables for prescriptions in each of the BNF chapters.

@@

# Discussion  
### Sociodemographic features
In this analysis of sociodemographic factors and past medical history we found a large number of factors to contribute to risk of severe COVID-19 disease.  Apart from the very steep rise in risk with age, severe disease is more likely in men, and those who are more socioeconomically disadvantaged.  Regarding sex differences we noted that the ratio of non-fatal cases who receive critical care to fatal cases is lower in females than in males in the 50-70 year age range.  This may reflect either an underlying divergence in case fatality, or that women are less likely than men to be transferred to critical care if they have a good chance of surviving without critical care.  The effect of matching on general practice which will partly match for socioeconomic disadvantage will reduce the precision of the effect of the socioeconomic indicator (SIMD) in our analysis but will not bias it in either direction.  Residence in a care home was associated with a 15-fold increased risk of severe COVID-19 in this age matched analysis and this was reduced to eight-fold adjusted for the listed conditions. This eight-fold risk likely reflects both the spread of the epidemic in care homes and some residual confounding by frailty.

In contrast to reports from the US and England [@the_opensafely_collaborative_opensafely_2020] we found no evidence of increased risk in non-White minorities. This suggests that ethnic differences reported in other countries are not genetically driven but reflect other social factors that differ between regions in their association with ethnicity.  We note that the proportion of health care workers that are of non-British nationality varies across the UK from [23% in London to 7% in Scotland](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/internationalmigration/articles/internationalmigrationandthehealthcareworkforce/2019-08-15).  In a recent report based on the UK Biobank cohort, almost half of the fourfold risk ratio for hospitalised COVID associated with Black ethnicity was accounted for by health care worker status, with other social variables attenuating the risk further[@niedzwiedz_ethnic_2020; @ho_modifiable_2020].  In the OpenSAFELY prospective cohort study using medical records in England compared to those recorded as White, Black and Asian people were at 1.7 fold and 1.6 fold risk of death respectively adjusted for comorbidities and socioeconomic status, However as acknowledged by the authors, data on health care worker status and other social factors would be needed to explore this further. We are in the process of obtaining more detailed data on health care worker status in Scotland.

### Co-morbidities
We have confirmed the increased risks associated with the listed conditions. These were based on the list of " moderate risk " conditions highlighted by the NHS and other agencies [@nhs_whos_2020].

The risk associated with these conditions various by age - for example the odds ratio associated with diabetes is much higher at younger ages. The 2.7 fold risk in all ages combined is commensurate with that reported in UK Biobank and in the OpenSAFELY studies.  We confirm the higher risk with asthma and chronic lung disease and liver disease reported in these and earlier studies.  Of note other heart disease is more strongly associated than ischaemic heart disease.  This category includes many conditions such as atrial fibrillation, cardiomyopathies and heart failure.  However overall 26% of cases did not have any of these conditions.  Among cases and controls without these conditions not surprisingly neoplasms were associated with severe COVID-19; we had omitted it from the pre-specified list as many people under current chemotherapy and radiotherapy are included in the shielding at present.  We have not focussed on evaluating the risk associated with conditions for which shielding is recommended, as the observed risk will depend on the adequacy of shielding.  A further systematic evaluation of past hospitalisation history did not reveal a sparse set of conditions underlying severe disease but rather very many associated diagnoses. We note however that past hospitalisation for influenza and respiratory tract infections is an important risk factor.

Media reports of apparently healthy young people succumbing to severe COVID-19 can lead to an impression that severe COVID-19 is a random event.  However we found that almost half of cases under 40 years had at least one of the listed conditions In those that did not have a listed condition having any hospital admission in the past five years and having any prescription in the past year was much more common among cases that in the background population of the same age. Thus the vast majority of younger persons succumbing to severe disease are not people with no prior health conditions.

## Possible drug effects
A striking finding of this study was the very strong association of severe COVID-19 with having encashed at least one prescription in the past year. This association was only partly explained by differing prescription levels associations with the listed conditions.  Partitioning of this association between BNF chapters, which represent broad indication-based drug classes, showed that the strongest association was with prescription of Chapter 1 drugs, prescribed for gastrointestinal conditions, which are not generally listed as risk factors for severe COVID-19 .  Also associated were those in the cardiovascular, nervous system and nutritional and blood chapters.  It is likely that most drug associations will simply reflect co-morbidities for which they are prescribed. However we note that there have been proposals of directly adverse effects of drugs such as angiotensin converting enzyme inhibitors [@fang_are_2020] although in a case control study from Italy differences in usage between cases with COVID-19 and controls were attributable to co-morbidities in a case control analysis [@mancia_reninangiotensinaldosterone_2020]

Nonetheless the strong association with encashed of at least one prescription warrants a more detailed investigation.  In an accompanying paper we explore these associations more fully since they require a detailed consideration of confounding by indication before and directly causal relationship can be asserted.
